Exploring the evolution of stellar rotation using Galactic kinematics

### 2020-04-24 paper/ms63.tex
adding citation to Ruprecht paper
small edits

\author{Elisabeth Newton}
\affiliation{Department of Physics and Astronomy, Dartmouth
College, Hanover, NH, USA}

@@

\author{Yuxi (Lucy) Lu}
\affiliation{Department of Astronomy, Columbia University, Manhattan, NY, USA}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

\author{Lauren Anderson}
\affiliation{Observatories of the Carnegie Institution for
Science, 813 Santa Barbara Street, Pasadena, CA 91101, USA}

@@

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a star’s
age from its rotation period \citep[\eg][]{skumanich1972, barnes2003,
barnes2007, mamajek2008, barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} confirm that rotational evolution is a highly complex
process.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
% In other words, convection zone depth cannot be the only variable that affects
% stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the Galaxy with a low out-of-plane, or vertical, velocity (\vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of older stars is observed to be larger
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
This is likely either a signature of dynamical heating, such as from
interactions with giant molecular clouds, spiral arms and the galactic bar
\citep[see][for a review of secular evolution in the MW]{sellwood2014}, or an
indication that stars formed dynamically ``hotter'' in the past
\citep[e.g.,][]{bird2013}.
In either case, the vertical velocity distribution is observed to depend
significantly on stellar age.
While the velocity of any individual star only provides a weak age constraint
\racomment{(if any at all)}, because its velocity depends on its current
position in its orbit, the velocity {\it dispersion} of a {\it population} of
stars indicates whether that population is old or young relative to other
populations.
% Vertical velocity dispersion is thought to increase monotonically with age.
In this work, we compare the velocity dispersions of populations of field
stars in the Galactic thin disk to ascertain which populations are older and
which younger, and draw conclusions about the rotational evolution of stars
based on their implied relative ages.

@@

Although {\it vertical} velocity, \vz, is \racomment{a well-established} age
proxy, it can only be calculated with full 6-dimensional position and velocity
information.
In fact, with full 6D phase space and an assumed Galactic potential, it is
possible to calculate the dynamically-invariant vertical {\it action}, which may
be an even better age indicator \citep{beane2018, ting2019}.
Unfortunately, most field stars with measured rotation periods do not have
radial velocity (RV) measurements because they are relatively faint \kepler\
targets ($\sim$12th-16th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb,
as a proxy for \vz.
The \kepler\ field is positioned at low galactic latitude
(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see
appendix) approximation to \vz.
Because we use \vb\ rather than \vz\, we do not calculate absolute kinematic
ages using a published age--velocity dispersion relation (AVR), calibrated with
vertical velocity.
In the future it may be possible to account for the differences between \vb\
and \vz, or marginalize over missing RV measurements and the \kepler\
selection function, in order to infer the absolute ages of populations of
stars.
Regardless of direction however, velocity dispersion is expected to
monotonically increase over time \citep[\eg][]{holmberg2009}, and can
therefore be used to {\it rank} populations of stars by age.

@@

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog of parallaxes, proper motions and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.
For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
\racomment{The calibration of this relation is described in Curtis \etal\ (2020,
in prep) and summarized in the Appendix of this paper.}

@@

The rotation periods of the dwarf stars in the \mct\ sample are shown on a
\gaia\ color-magnitude diagram (CMD) in the top panel of figure
\ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2008, newton2017, kiman2019}, and are therefore more likely
to have measurable rotation periods at old ages, the latter scenario seems
likely.
However, it is also possible that the rotation periods of the oldest early M
dwarfs are so long that they are not measurable with Kepler data.
\racomment{Ground-based rotation period measurements of early M dwarfs
(spectral types earlier than $\sim$ M2.5, \teff $\gtrsim$3500 K) indicate that
there is an upper limit to their rotation periods of around 80 days
\citep{newton2016, newton2018}, which is longer than the longest rotation
periods measured for the early M dwarfs in the \mct\ sample (around 50 days).}
The apparent lack of old gyro-ages for M dwarfs in figure
\ref{fig:age_gradient} may be caused by a combination of ages being
underestimated by a poorly calibrated model, and rotation period detection
bias.
The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
Inaccuracies at low masses are a typical feature of empirically calibrated
gyrochronology models since there are no (or at least very few) old M dwarfs
with rotation periods and the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
(around 3,000 stars), stars fainter than 16th magnitude \racomment{in \gaia\
G-band} (200 stars), stars with absolute \vb\ uncertainties greater than 1
\kms\ (1000 stars), and stars with galactic latitudes greater than 15\degrees\
(5500 stars, justification provided in the appendix) were removed from the
sample.
Finally, we removed almost 2000 stars with rotation periods shorter than the
main population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
\racomment{These rapid rotators are probably either very young stars, or
synchronized binaries (see section \ref{sec:gap}).} After these cuts, around
13,000 stars out of the original $\sim$34,000 were included in the sample.

@@

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citet{spada2019} model shown in white.
These gyrochrones are \racomment{plotted for the same ages described above.}
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduce the apparent
flattening and inversion in the rotation period-\teff/mass relations.

@@

The velocity dispersions of stars in the \mct\ sample, shown in figure
\ref{fig:vplot}, provide the following picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `C' sequence to the `I'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010, matt2012, matt2015}.
% This aligns with the current assumptions about stellar spin-down, dynamo
% theory, and the gyrochronology paradigm that has been in place for decades.
% \citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, barnes2010}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages \racomment{compared to higher-mass stars}, but their cores are more
tightly coupled to their envelopes, allowing angular momentum transport
between the two layers.
Angular momentum surfaces \racomment{from the core} and prevents the stellar
envelopes from spinning-down rapidly, and this effect is strongest for late
K-dwarfs with effective temperatures of $\sim$4000-4500K and masses
$\sim$0.5-0.7 M$_\odot$.

@@

The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.
Most semi-empirical spin-down models predict that, once the rotation periods of
stars converge to a tight sequence, they approximately spin down with a common
braking index \citep[e.g., Fig. 5 in][]{gallet2015}.
While some semi-empirical models do not predict that stars have fully converged
by the age of Praesepe, the observations suggest that they have
\citep{douglas2019}, in which case, projecting the Praesepe sequence forward
with a single braking index that is common to all stars and constant in time
is appropriate.
However, observations of low-mass stars in older clusters \citep[\eg, NGC
6811, NGC 752, and Ruprecht 147][Curtis \etal, submitted]{curtis2019,
agueros2018} demonstrate that the assumptions adopted in those models must be
invalid, or inaccurate in some quantitative way (for example, the
core-envelope coupling timescale is inaccurately calibrated).

There are many other empirical gyrochronology models we could have chosen to
compare with our data.
For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015} relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
0.5), which applies to stars of all masses and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data here.
The \citet{barnes2010} model has a similar period-\teff\ relation to the
Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
Praesepe roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
in its current form does not have the flexibility to capture the mass and
time-variable rotational evolution revealed by the data explored here.

@@

It is not surprising that the \citet{angus2019} model, or other gyrochronology
models do not reproduce the data, since tensions with
\citet{skumanich1972}-type spin-down have been revealed whenever new rotation
periods of benchmark stars have become available.
With the influx of new stellar rotation periods provided by \kepler/\ktwo\ and
\tess, our understanding of spin-down is changing rapidly.
New data are revealing flaws in not just the calibration of but also the
functional forms of old models.
We have shown that velocity dispersion data provides support for more
sophisticated models incorporating additional physical effects \citep[\eg,
core-envelope coupling][]{spada2019} over simple empirical models.
Alternatively, more flexible empirical frameworks, like a Gaussian process
model, could capture everything we now know about stellar spin-down.
}

@@

The top panel of figure \ref{fig:vplot} suggests that the \citet{angus2019},
Praesepe-based gyrochronology model is valid below the gap but not above.
Gyrochrones follow lines of constant velocity dispersion below the gap, but
{\it cross} lines of constant velocity dispersion above the gap.
This phenomenon is robust to the choice of bin size and position.
Although we do not provide an in-depth analysis here (and more data may be
needed to confirm a connection) these data suggest that the gap may indeed
separate a young regime where stellar cores are decoupled from their envelopes
from an old regime where these layers are more tightly coupled.
If so, this could indicate that \racomment{these phenomena are related, \ie\
the process that is responsible for changing the shape of gyrochrones in
rotation-\teff\ space is related to the process that produces the gap.
% Of course this may just be a coincidence.
% Alternatively, the fact that the gap is located near a changing stellar
% evolution regime may be a coincidence.
}

@@

We thank the anonymous referee for their comments which greatly improved the
manuscript.
We would like to thank Suzanne Aigrain for providing thoughtful insight that
improved this paper.
This work was partly developed at the 2019 KITP conference `Better stars,
better planets'.
This research was supported in part by the National Science Foundation under
Grant No. NSF PHY-1748958.
Parts of this project are based on ideas explored at the Gaia sprints at the
Flatiron Institute in New York City, 2016 and MPIA, Heidelberg, 2017.
This work made use of the gaia-kepler.fun crossmatch database created by Megan
Bedell.
T.A.B. acknowledges support by a NASA FINESST award (80NSSC19K1424).

>>>>

\author[0000-0003-4150-841X]{Elisabeth Newton}
\affiliation{Department of Physics and Astronomy, Dartmouth
College, Hanover, NH, USA}

@@

\author[0000-0003-4769-3273]{Yuxi (Lucy) Lu}
\affiliation{Department of Astronomy, Columbia University, Manhattan, NY, USA}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

\author[0000-0001-5725-9329]{Lauren Anderson}
\affiliation{Observatories of the Carnegie Institution for
Science, 813 Santa Barbara Street, Pasadena, CA 91101, USA}

@@

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a star’s
age from its rotation period \citep[\eg][]{skumanich1972, barnes2003,
barnes2007, mamajek2008, barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} confirm that rotational evolution is a highly complex
process.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
\racomment{
    In addition, new rotation period measurements for low-mass stars in
the 2.7-Gyr-old cluster Ruprecht 147 show that its slow-rotator sequence is
flat compared to younger clusters like Praesepe (Curtis \etal\ 2020, submitted
to AAS Journals), and departures from classical Skumanich-like spin-down were
also noted in observations of the $\sim$1.3 Gyr NGC 752 open cluster
\citep{agueros2018}.
}
% In other words, convection zone depth cannot be the only variable that
% affects stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce the flattened
period--color relation
\\racomment{
    seen in these clusters \citep{spada2019}.
}
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the Galaxy with a low out-of-plane, or vertical, velocity (\vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of older stars is observed to be larger
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
This is likely either a signature of dynamical heating, such as from
interactions with giant molecular clouds, spiral arms and the galactic bar
\citep[see][for a review of secular evolution in the MW]{sellwood2014}, or an
indication that stars formed dynamically ``hotter'' in the past
\citep[e.g.,][]{bird2013}.
In either case, the vertical velocity distribution is observed to depend
significantly on stellar age.
While the velocity of any individual star only provides a weak age constraint
\racomment{
    (if any at all)
}, because its velocity depends on its current
position in its orbit, the velocity {\it dispersion} of a {\it population} of
stars indicates whether that population is old or young relative to other
populations.
% Vertical velocity dispersion is thought to increase monotonically with age.
In this work, we compare the velocity dispersions of populations of field
stars in the Galactic thin disk to ascertain which populations are older and
which younger, and draw conclusions about the rotational evolution of stars
based on their implied relative ages.

@@

Although {\it vertical} velocity, \vz, is \racomment{
    a well-established
}
age proxy, it can only be calculated with full 6-dimensional position and
velocity information.
In fact, with full 6D phase space and an assumed Galactic potential, it is
possible to calculate the dynamically-invariant vertical {\it action}, which may
be an even better age indicator \citep{beane2018, ting2019}.
Unfortunately, most field stars with measured rotation periods do not have
radial velocity (RV) measurements because they are relatively faint \kepler\
targets ($\sim$12th-16th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb,
as a proxy for \vz.
The \kepler\ field is positioned at low galactic latitude
(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect, see
appendix) approximation to \vz.
Because we use \vb\ rather than \vz\, we do not calculate absolute kinematic
ages using a published age--velocity dispersion relation (AVR), calibrated with
vertical velocity.
In the future it may be possible to account for the differences between \vb\
and \vz, or marginalize over missing RV measurements and the \kepler\
selection function, in order to infer the absolute ages of populations of
stars.
Regardless of direction however, velocity dispersion is expected to
monotonically increase over time \citep[\eg][]{holmberg2009}, and can
therefore be used to {\it rank} populations of stars by age.

@@

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog of parallaxes, proper motions and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.
For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
\racomment{
    The calibration of this relation is described in Curtis \etal\ (2020,
in prep) and summarized in the Appendix of this paper.
}

@@

The rotation periods of the dwarf stars in the \mct\ sample are shown on a
\gaia\ color-magnitude diagram (CMD) in the top panel of figure
\ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2008, newton2017, kiman2019}, and are therefore more likely
to have measurable rotation periods at old ages, the latter scenario seems
likely.
However, it is also possible that the rotation periods of the oldest early M
dwarfs are so long that they are not measurable with Kepler data.
\racomment{
    Ground-based rotation period measurements of early M dwarfs
(spectral types earlier than $\sim$ M2.5, \teff\ $\gtrsim$ 3500 K) indicate
there is a $\sim$ 80 day upper limit to their rotation periods
\citep{newton2016, newton2018}, which is longer than the longest rotation
periods measured for the early M dwarfs in the \mct\ sample (around 50 days).
}

The apparent lack of old gyro-ages for M dwarfs in figure
\ref{fig:age_gradient} may be caused by a combination of ages being
underestimated by a poorly calibrated model, and rotation period detection
bias.
The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
Inaccuracies at low masses are a typical feature of empirically calibrated
gyrochronology models since there are no (or at least very few) old M dwarfs
with rotation periods and the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
(around 3,000 stars), stars fainter than 16th magnitude \racomment{
    in \gaia\ G-band
}
(200 stars), stars with absolute \vb\ uncertainties greater than 1
\kms\ (1000 stars), and stars with galactic latitudes greater than 15\degrees\
(5500 stars, justification provided in the appendix) were removed from the
sample.
Finally, we removed almost 2000 stars with rotation periods shorter than the
main population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
\racomment{
    These rapid rotators are probably either very young stars, or
synchronized binaries (see section \ref{sec:gap}).
}
After these cuts, around
13,000 stars out of the original $\sim$34,000 were included in the sample.

@@

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citet{spada2019} model shown in white.
These gyrochrones are \racomment{
    plotted for the same ages described above.
}
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduce the apparent
flattening and inversion in the rotation period-\teff/mass relations.

@@

The velocity dispersions of stars in the \mct\ sample, shown in figure
\ref{fig:vplot}, provide the following picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `C' sequence to the `I'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010, matt2012, matt2015}.
% This aligns with the current assumptions about stellar spin-down, dynamo
% theory, and the gyrochronology paradigm that has been in place for decades.
% \citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, barnes2010}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages \racomment{
    compared to higher-mass stars,
}
but their cores are more
tightly coupled to their envelopes, allowing angular momentum transport
between the two layers.
Angular momentum surfaces \racomment{
    from the core
}
and prevents the stellar
envelopes from spinning-down rapidly, and this effect is strongest for late
K-dwarfs with effective temperatures of $\sim$4000-4500K and masses
$\sim$0.5-0.7 M$_\odot$.

@@

The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.
Most semi-empirical spin-down models predict that, once the rotation periods of
stars converge to a tight sequence, they approximately spin down with a common
braking index \citep[e.g., Fig. 5 in][]{gallet2015}.
So, if stellar rotation periods have fully converged by the age of Praesepe,
as the observations suggest they have \citep{douglas2019}, it is appropriate
to project the Praesepe sequence forward in time with a single braking index
that is common to all stars and constant in time.
% While some semi-empirical models do not predict that stars have
% by the age of Praesepe, the observations suggest that they have
% \citep{douglas2019}, in which case, projecting the Praesepe sequence forward
% with a single braking index that is common to all stars and constant in time
% is appropriate.
However, observations of low-mass stars in older clusters \citep[\eg, NGC
6811, NGC 752, and Ruprecht 147][Curtis \etal\ 2020, submitted]{curtis2019,
agueros2018} now demonstrate that the assumptions adopted in the
semi-empirical models which support a common braking index must be invalid, or
inaccurate in some quantitative way (for example, the core-envelope coupling
timescale is inaccurately calibrated).

There are many other empirical gyrochronology models we could have chosen to
compare with our data.
For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015} relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$
\citep[$\approx$ 0.5,][]{skumanich1972}, which applies to stars of all masses
and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data here.
The \citet{barnes2010} model has a similar period-\teff\ relation to the
Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
Praesepe roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
in its current form does not have the flexibility to capture the mass and
time-variable rotational evolution revealed by the data explored here.

@@

It is not surprising that the \citet{angus2019} model, or other gyrochronology
models do not reproduce the data, since tensions with Skumanich-like spin-down
have been revealed whenever new rotation periods of benchmark stars have
become available \citep{angus2015, vansaders2016, metcalfe2019, agueros2018,
curtis2019}.
With the influx of new stellar rotation periods provided by \kepler/\ktwo,
\tess, and ground-based facilities, our understanding of spin-down is changing
rapidly.
New data are revealing flaws in not just the calibration of but also the
functional forms of old models.
We have shown that velocity dispersion data provides support for more
sophisticated models that incorporate additional physical effects \citep[\eg,
core-envelope coupling][]{spada2019} over the simple empirical models used
today.
Alternatively, more flexible empirical frameworks, like a Gaussian process
model, could capture everything we now know about stellar spin-down.
}

@@

The top panel of figure \ref{fig:vplot} suggests that the \citet{angus2019},
Praesepe-based gyrochronology model is valid below the gap but not above.
Gyrochrones follow lines of constant velocity dispersion below the gap, but
{\it cross} lines of constant velocity dispersion above the gap.
This phenomenon is robust to the choice of bin size and position.
Although we do not provide an in-depth analysis here (and more data may be
needed to confirm a connection) these data suggest that the gap may indeed
separate a young regime where stellar cores are decoupled from their envelopes
from an old regime where these layers are more tightly coupled.
If so, this could indicate that
\racomment{
    these phenomena are related, \ie\
the process that is responsible for changing the shape of gyrochrones in
rotation-\teff\ space is related to the process that produces the gap.
% Of course this may just be a coincidence.
% Alternatively, the fact that the gap is located near a changing stellar
% evolution regime may be a coincidence.
}

@@

We thank the anonymous referee for their comments which greatly improved this
manuscript.
We also would like to thank Suzanne Aigrain for providing thoughtful insight
that improved the paper.
This work was partly developed at the 2019 KITP conference `Better stars,
better planets'.
This research was supported in part by the National Science Foundation under
Grant No. NSF PHY-1748958.
Parts of this project are based on ideas explored at the Gaia sprints at the
Flatiron Institute in New York City, 2016 and MPIA, Heidelberg, 2017.
This work made use of the gaia-kepler.fun crossmatch database created by Megan
Bedell.
T.A.B. acknowledges support by a NASA FINESST award (80NSSC19K1424).
